---
title: "Render manuscript figures (R tier)"
output: html_document
---

```{r setup, include=FALSE}
find_v2_root <- function() {
  d <- normalizePath(getwd(), mustWork = TRUE)
  while (nchar(d) > 1) {
    if (file.exists(file.path(d, "config.py")) &&
        dir.exists(file.path(d, "R"))) return(d)
    parent <- dirname(d)
    if (parent == d) break
    d <- parent
  }
  stop("Could not find v2 root from ", getwd())
}

V2_ROOT <- find_v2_root()
knitr::opts_knit$set(root.dir = V2_ROOT)
```

Runs the `R/plot_figure_*.R` scripts in this notebook's **R kernel**. Each
script reads the CSVs written by the Python prep tier and builds every panel as a standalone
ggplot object, saving each one individually (both PNG and PDF) — there is no composed
multi-panel figure.

Metric-aware figures render twice: once using mean time-dependent AUC(t), the primary metric,
and once using Harrell's C-index. Metric-invariant figures render once.

- **Prep tier**: [06_generate_figure_data.ipynb](06_generate_figure_data.ipynb) (Python kernel) — run first.
- **One-time R package bootstrap** (any host with R):

  ```
  Rscript v2/R/install_packages.R
  ```

Each plot script is `sys.source()`-ed in its own environment so per-script ggplot objects
(`p2a`, `p3b`, …) don't leak between cells. v2 R scripts always assume the working directory
is `v2/` (they `source("R/figure_utils.R")` with a relative path), so this document sets the
R working directory to the v2 root (via `knitr::opts_knit$set(root.dir = ...)`, which — unlike a
bare `setwd()` in a chunk — persists across all chunks) before sourcing anything.

```{r find-root}
R_DIR <- file.path(V2_ROOT, "R")

cat("v2 root:   ", V2_ROOT, "\n", sep = "")
cat("R scripts: ", R_DIR, "\n", sep = "")
cat("R version: ", R.version.string, "\n", sep = "")
```

```{r render-plots}
METRIC_INVARIANT_SCRIPTS <- c(
  "plot_figure_0.R",
  "plot_figure_1.R",
  "plot_figure_4.R",
  "plot_figure_4_supp.R",
  "plot_figure_5.R"
)

METRIC_SCRIPTS <- c(
  "plot_figure_2.R",
  "plot_figure_2_supp.R",
  "plot_figure_2_supp_events.R",
  "plot_figure_2_supp_anchor.R",
  "plot_figure_3.R"
)

run_plot <- function(script, metric = NULL) {
  if (is.null(metric)) Sys.unsetenv("MANUSCRIPT_METRIC") else Sys.setenv(MANUSCRIPT_METRIC = metric)
  cat("\n=== ", script, if (is.null(metric)) "" else paste0(" [", metric, "]"), " ===\n", sep = "")
  env <- new.env(parent = globalenv())
  sys.source(file.path(R_DIR, script), envir = env)
}

for (s in METRIC_INVARIANT_SCRIPTS) run_plot(s)
for (metric in c("auc", "cindex")) {
  for (s in METRIC_SCRIPTS) run_plot(s, metric)
}

Sys.unsetenv("MANUSCRIPT_METRIC")

cat("\nDone.\n")
cat("Panel PNGs: $CLINICAL_FIGURES_OUT/png/<figureN>/ (or /data/.../manuscript_figures/png/<figureN>/)\n")
cat("Panel PDFs: $CLINICAL_FIGURES_OUT/pdf/<figureN>/ (or /data/.../manuscript_figures/pdf/<figureN>/)\n")
```

Render with:

```
Rscript -e 'rmarkdown::render("v2/notebooks/07_render_figures.Rmd")'
```
