---
title: "Semantic-search figures"
subtitle: "Patient embedding PC space, clinical associations, and XGBoost recovery"
output:
  html_document:
    toc: true
    toc_float: true
    code_folding: hide
params:
  space: "concat"
  window: "alltime"
  max_pcs: 20
---

This is the figures/reporting tier for the semantic-search arm. It reads the
artifacts produced by stages 2--4; it does not fit PCA models or retrain
classifiers. By default it reports the `concat/alltime` analysis. Change the
parameters at render time to inspect another completed space/window pair.

The `alltime` window is descriptive: notes can contain information written
after treatment or near an outcome. Consequently, these panels measure
retrospective clinical-content recovery, rather than prospective prediction or
independent validation.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 6, dpi = 160)

find_repo_root <- function(start = getwd()) {
  candidate <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(candidate, "config.py")) &&
        dir.exists(file.path(candidate, "semantic_search"))) return(candidate)
    parent <- dirname(candidate)
    if (identical(parent, candidate)) stop("Could not find repository root from ", start)
    candidate <- parent
  }
}

REPO_ROOT <- find_repo_root()
knitr::opts_knit$set(root.dir = REPO_ROOT)
setwd(REPO_ROOT)

suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(ggplot2)
  library(readr)
  library(tidyr)
})
source("R/config.R")

SPACE <- params$space
WINDOW <- params$window
MAX_PCS <- as.integer(params$max_pcs)
SEMANTIC_SEARCH_PATH <- file.path(DATA_PATH, "semantic_search")
PCS_DIR <- file.path(SEMANTIC_SEARCH_PATH, "pcs")
RESULTS_DIR <- file.path(SEMANTIC_SEARCH_PATH, "results")
PREDICTIONS_DIR <- file.path(SEMANTIC_SEARCH_PATH, "predictions")
FIGURE_DIR <- Sys.getenv(
  "SEMANTIC_SEARCH_FIGURES_OUT",
  unset = file.path(FIGURE_OUT_DIR, "semantic_search", paste(SPACE, WINDOW, sep = "_"))
)
dir.create(FIGURE_DIR, recursive = TRUE, showWarnings = FALSE)

score_file <- file.path(PCS_DIR, sprintf("%s_%s_scores.parquet", SPACE, WINDOW))
loading_file <- file.path(PCS_DIR, sprintf("%s_%s_loadings.parquet", SPACE, WINDOW))
variance_file <- file.path(RESULTS_DIR, "pc_explained_variance.csv")
association_file <- file.path(RESULTS_DIR, "pc_clinical_associations.csv")

need_file <- function(path, stage) {
  if (!file.exists(path)) stop(stage, " artifact not found: ", path,
                              "\nRun the corresponding semantic_search stage first.")
}
read_csv_if_present <- function(path) {
  if (!file.exists(path)) return(NULL)
  readr::read_csv(path, show_col_types = FALSE)
}
save_panel <- function(plot, name, width = 10, height = 6) {
  ggplot2::ggsave(file.path(FIGURE_DIR, paste0(name, ".png")), plot,
                  width = width, height = height, dpi = 300)
  ggplot2::ggsave(file.path(FIGURE_DIR, paste0(name, ".pdf")), plot,
                  width = width, height = height)
  plot
}
pc_number <- function(x) as.integer(sub("^PC", "", as.character(x)))
display_target <- function(x) {
  dplyr::recode(x,
    cancer_type = "Cancer type", stage = "Cancer stage",
    first_treatment = "First treatment type", prostate_subtype = "Prostate phenotype",
    .default = x
  )
}
theme_report <- theme_minimal(base_size = 11) +
  theme(panel.grid.minor = element_blank(),
        plot.title = element_text(face = "bold"),
        legend.position = "bottom")
```

```{r inputs}
need_file(score_file, "PC score")
need_file(loading_file, "PC loading")
need_file(variance_file, "Explained-variance")

scores <- as.data.frame(arrow::read_parquet(score_file))
loadings <- as.data.frame(arrow::read_parquet(loading_file))
variance <- readr::read_csv(variance_file, show_col_types = FALSE) %>%
  filter(space == SPACE, window == WINDOW) %>%
  mutate(component = as.integer(component)) %>% arrange(component)
association_raw <- read_csv_if_present(association_file)
associations <- if (is.null(association_raw)) tibble() else association_raw %>%
  filter(space == SPACE, window == WINDOW) %>%
  mutate(component = as.integer(component), fdr = as.numeric(fdr),
         p = as.numeric(p), effect = as.numeric(effect))
coverage_raw <- read_csv_if_present(file.path(RESULTS_DIR, "pc_join_coverage.csv"))
association_coverage <- if (is.null(coverage_raw)) tibble() else coverage_raw %>%
  filter(space == SPACE, window == WINDOW)
summary_raw <- read_csv_if_present(file.path(RESULTS_DIR, "prediction_metrics_summary.csv"))
metrics_summary <- if (is.null(summary_raw)) tibble() else summary_raw %>%
  filter(space == SPACE, window == WINDOW, model == "xgboost")
folds_raw <- read_csv_if_present(file.path(RESULTS_DIR, "prediction_metrics_folds.csv"))
metrics_folds <- if (is.null(folds_raw)) tibble() else folds_raw %>%
  filter(space == SPACE, window == WINDOW, model == "xgboost")
class_raw <- read_csv_if_present(file.path(RESULTS_DIR, "prediction_metrics_by_class.csv"))
metrics_class <- if (is.null(class_raw)) tibble() else class_raw %>%
  filter(space == SPACE, window == WINDOW, model == "xgboost")

cat("PC scores: ", nrow(scores), " patients; ", ncol(scores) - 1, " PCs\n", sep = "")
cat("Associations: ", nrow(associations), " tests\n", sep = "")
cat("XGBoost runs: ", nrow(metrics_summary), " target(s)\n", sep = "")
cat("Figure directory: ", FIGURE_DIR, "\n", sep = "")
```

## PC space

### Explained variance

```{r scree-plot}
if (nrow(variance) == 0) {
  cat("No variance rows found for this space/window pair.\n")
} else {
  scree <- ggplot(variance, aes(component, explained_variance_ratio)) +
    geom_col(fill = "#3B6FB6") +
    geom_line(aes(y = cumulative_explained_variance_ratio), colour = "#C84E4E",
              linewidth = 0.8) +
    geom_point(aes(y = cumulative_explained_variance_ratio), colour = "#C84E4E", size = 1.5) +
    scale_y_continuous(labels = function(x) paste0(round(100 * x), "%"),
                       sec.axis = sec_axis(~ ., name = "Cumulative explained variance",
                                           labels = function(x) paste0(round(100 * x), "%"))) +
    labs(title = "PCA scree plot", x = "Principal component",
         y = "Explained variance") + theme_report
  save_panel(scree, "01_scree")
}
```

### PC1--PC2 score space

Each point is a patient. Hexagonal binning keeps the display legible for a
large cohort and makes density structure visible without overplotting.

```{r pc-space}
if (!all(c("PC1", "PC2") %in% names(scores))) {
  cat("At least PC1 and PC2 are required for this panel.\n")
} else {
  pc_space <- ggplot(scores, aes(PC1, PC2)) +
    geom_bin2d(bins = 70) +
    scale_fill_viridis_c(name = "Patients", trans = "sqrt") +
    coord_equal() +
    labs(title = "Patient embedding space: PC1 vs PC2",
         subtitle = sprintf("%s / %s; n = %s", SPACE, WINDOW, format(nrow(scores), big.mark = ","))) +
    theme_report
  save_panel(pc_space, "02_pc1_pc2_space")
}
```

### Distribution of leading PC scores

```{r pc-score-distributions}
score_pcs <- names(scores)[grepl("^PC[0-9]+$", names(scores))]
score_pcs <- score_pcs[order(pc_number(score_pcs))]
score_pcs <- head(score_pcs, min(12, length(score_pcs)))
if (length(score_pcs) == 0) {
  cat("No PC columns found in the score artifact.\n")
} else {
  score_long <- scores %>%
    select(all_of(score_pcs)) %>%
    pivot_longer(everything(), names_to = "pc", values_to = "score") %>%
    mutate(pc = factor(pc, levels = score_pcs))
  score_dist <- ggplot(score_long, aes(score)) +
    geom_histogram(bins = 50, fill = "#3B6FB6", colour = "white", linewidth = 0.15) +
    facet_wrap(~ pc, scales = "free", ncol = 4) +
    labs(title = "Leading PC score distributions", x = "PC score", y = "Patients") + theme_report
  save_panel(score_dist, "03_pc_score_distributions", width = 12, height = 8)
}
```

### Contribution of note types to leading PCs

The bars show each note type's share of the squared PCA loading. Since all
three blocks have the same number of embedding dimensions, this provides a
compact view of whether a component is predominantly carried by one note type.

```{r loading-by-note-type}
loading_share <- loadings %>%
  mutate(component = as.integer(component)) %>%
  filter(component <= MAX_PCS) %>%
  group_by(component, pc, note_type) %>%
  summarise(squared_loading = sum(loading ^ 2), .groups = "drop") %>%
  group_by(component, pc) %>%
  mutate(share = squared_loading / sum(squared_loading)) %>% ungroup() %>%
  mutate(pc = factor(pc, levels = paste0("PC", sort(unique(component)))))
if (nrow(loading_share) == 0) {
  cat("No loading rows found for the requested PCs.\n")
} else {
  loading_plot <- ggplot(loading_share, aes(pc, share, fill = note_type)) +
    geom_col() +
    scale_y_continuous(labels = function(x) paste0(round(100 * x), "%")) +
    scale_fill_manual(values = c(Clinician = "#3B6FB6", Imaging = "#E39D3A", Pathology = "#5B9E77")) +
    labs(title = "Note-type contribution to PC loadings", x = "Principal component",
         y = "Share of squared loading", fill = "Note type") + theme_report +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  save_panel(loading_plot, "04_note_type_loading_share", width = 12, height = 6)
}
```

## Associations with PC scores

Associations use Spearman correlation for continuous variables, Kruskal--Wallis
with epsilon-squared for categorical variables, and Cox proportional hazards
models for overall survival. FDR is adjusted across the full PC screen within
each clinical family, so association effect estimates are intentionally kept
separate by test type below.

### Association discovery map

For every clinical family and PC, the map displays the variable with the
smallest within-family FDR. A blank/white tile means no finite adjusted p-value
was available, not evidence of no association.

```{r association-map}
if (nrow(associations) == 0) {
  cat("PC association results are absent. Run semantic_search.correlate_pcs first.\n")
} else {
  association_map <- associations %>%
    filter(component <= MAX_PCS, is.finite(fdr)) %>%
    group_by(family, component, pc) %>%
    slice_min(fdr, n = 1, with_ties = FALSE) %>% ungroup() %>%
    mutate(signal = -log10(pmax(fdr, .Machine$double.xmin)),
           label = ifelse(nchar(variable) > 22, paste0(substr(variable, 1, 21), "…"), variable),
           pc = factor(pc, levels = paste0("PC", sort(unique(component)))))
  if (nrow(association_map) == 0) {
    cat("No finite FDR values available for the requested PCs.\n")
  } else {
    map_plot <- ggplot(association_map, aes(pc, family, fill = signal)) +
      geom_tile(colour = "white", linewidth = 0.25) +
      geom_text(aes(label = label), size = 2.5) +
      scale_fill_viridis_c(name = expression(-log[10](FDR)), na.value = "grey95") +
      labs(title = "Most significant PC association within each family",
           subtitle = "Tile labels name the best-associated clinical variable", x = "Principal component", y = NULL) +
      theme_report + theme(axis.text.x = element_text(angle = 45, hjust = 1))
    save_panel(map_plot, "05_association_discovery_map", width = 14, height = 7)
  }
}
```

### Clinical-family join coverage

Low join coverage means a family is evaluated on a selected subset of the PC
cohort. Inspect this before treating absence of an association as informative.

```{r association-coverage}
if (nrow(association_coverage) > 0) {
  knitr::kable(
    association_coverage %>%
      select(family, n_pc_patients, n_matched, coverage, below_threshold) %>%
      arrange(coverage) %>%
      mutate(coverage = scales::percent(coverage, accuracy = 0.1)),
    caption = "Patient overlap between PC scores and each clinical family."
  )
}
```

### Effect-size overview by test type

Spearman rho is signed; epsilon-squared is non-negative; Cox effects are hazard
ratios per one-PC standard deviation. They should not be compared on a common
numeric scale.

```{r association-effects}
if (nrow(associations) > 0) {
  top_effects <- associations %>%
    filter(component <= MAX_PCS, is.finite(fdr), is.finite(effect)) %>%
    group_by(family, test) %>% slice_min(fdr, n = 12, with_ties = FALSE) %>% ungroup() %>%
    mutate(label = paste0(pc, ": ", variable),
           label = ifelse(nchar(label) > 34, paste0(substr(label, 1, 33), "…"), label),
           label = reorder(label, effect),
           null_effect = ifelse(effect_name == "hazard_ratio_per_sd", 1, 0))
  if (nrow(top_effects) > 0) {
    effect_plot <- ggplot(top_effects, aes(label, effect, colour = fdr < 0.05)) +
      geom_hline(aes(yintercept = null_effect), colour = "grey75") +
      geom_point(size = 2) + coord_flip() +
      facet_wrap(~ effect_name, scales = "free_x") +
      scale_colour_manual(values = c(`TRUE` = "#C84E4E", `FALSE` = "#555555"),
                          name = "FDR < 0.05") +
      labs(title = "Leading clinical associations by compatible effect measure",
           x = NULL, y = "Effect estimate") + theme_report
    save_panel(effect_plot, "06_association_effects", width = 13, height = 9)
  }
}
```

### FDR-significant association table

```{r significant-association-table}
if (nrow(associations) > 0) {
  significant_associations <- associations %>%
    filter(is.finite(fdr), fdr < 0.05) %>% arrange(fdr, desc(abs(effect))) %>%
    select(pc, family, variable, test, effect_name, effect, fdr, n) %>% head(50)
  if (nrow(significant_associations) == 0) {
    cat("No FDR-significant PC associations in this completed result set.\n")
  } else {
    knitr::kable(significant_associations, digits = 3,
                 caption = "Up to 50 associations with FDR < 0.05, ordered by adjusted p-value.")
  }
}
```

## XGBoost clinical-label recovery

Each value below comes from out-of-fold predictions of the nested-CV XGBoost
workflow. The embedding preprocessing and blockwise PCA are fit inside each
training partition. These are therefore valid internal out-of-fold recovery
metrics, but alltime labels may still be directly documented in the notes.

### Aggregate out-of-fold metrics

```{r xgboost-summary}
if (nrow(metrics_summary) == 0) {
  cat("No XGBoost summary found for this space/window. Run semantic_search.train_prediction_models first.\n")
} else {
  metric_columns <- c(
    "balanced_accuracy_pooled_oof", "macro_f1_pooled_oof",
    "macro_ovr_auc_pooled_oof", "macro_average_precision_pooled_oof"
  )
  available_metrics <- intersect(metric_columns, names(metrics_summary))
  metric_labels <- c(
    balanced_accuracy_pooled_oof = "Balanced accuracy",
    macro_f1_pooled_oof = "Macro F1",
    macro_ovr_auc_pooled_oof = "Macro one-vs-rest AUROC",
    macro_average_precision_pooled_oof = "Macro average precision"
  )
  summary_long <- metrics_summary %>%
    select(target, n_patients, all_of(available_metrics)) %>%
    pivot_longer(all_of(available_metrics), names_to = "metric", values_to = "value") %>%
    mutate(target = display_target(target), metric = unname(metric_labels[metric]))
  metric_plot <- ggplot(summary_long, aes(target, value, fill = metric)) +
    geom_col(position = position_dodge(width = 0.8), width = 0.72) +
    geom_text(aes(label = sprintf("%.2f", value)), position = position_dodge(width = 0.8),
              vjust = -0.25, size = 3) +
    scale_y_continuous(limits = c(0, 1.08)) +
    labs(title = "Pooled out-of-fold XGBoost performance", x = NULL, y = "Metric value", fill = NULL) +
    theme_report + theme(axis.text.x = element_text(angle = 25, hjust = 1))
  save_panel(metric_plot, "07_xgboost_pooled_metrics", width = 12, height = 6)

  knitr::kable(metrics_summary %>%
    select(target, n_patients, n_classes, all_of(available_metrics)) %>%
    mutate(target = display_target(target)), digits = 3,
    caption = "Pooled out-of-fold performance. Each patient contributes exactly one outer-fold prediction.")
}
```

### Outer-fold stability

```{r xgboost-fold-metrics}
if (nrow(metrics_folds) > 0) {
  fold_columns <- intersect(c("balanced_accuracy", "macro_f1", "macro_ovr_auc", "macro_average_precision"),
                            names(metrics_folds))
  fold_long <- metrics_folds %>%
    select(target, fold, all_of(fold_columns)) %>%
    pivot_longer(all_of(fold_columns), names_to = "metric", values_to = "value") %>%
    mutate(target = display_target(target),
           metric = gsub("_", " ", metric))
  fold_plot <- ggplot(fold_long, aes(target, value)) +
    geom_boxplot(fill = "#9AB9DC", outlier.shape = NA) +
    geom_jitter(width = 0.1, height = 0, size = 1.8, colour = "#244A7B") +
    facet_wrap(~ metric, scales = "free_y") +
    labs(title = "Outer-fold metric stability", x = NULL, y = "Metric value") + theme_report +
    theme(axis.text.x = element_text(angle = 25, hjust = 1))
  save_panel(fold_plot, "08_xgboost_outer_fold_stability", width = 13, height = 7)
}
```

### Per-class recovery

```{r xgboost-class-metrics}
if (nrow(metrics_class) > 0) {
  class_plot_data <- metrics_class %>%
    select(target, class, n, recall, f1, ovr_auc, average_precision) %>%
    pivot_longer(c(recall, f1, ovr_auc, average_precision), names_to = "metric", values_to = "value") %>%
    mutate(target = display_target(target),
           row = paste(target, class, sep = " — "),
           row = reorder(row, n))
  class_plot <- ggplot(class_plot_data, aes(metric, row, fill = value)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = ifelse(is.finite(value), sprintf("%.2f", value), "NA")), size = 3) +
    scale_fill_viridis_c(limits = c(0, 1), na.value = "grey90", name = "Metric") +
    labs(title = "Per-class out-of-fold performance", x = NULL, y = NULL) + theme_report
  save_panel(class_plot, "09_xgboost_per_class_metrics", width = 11, height = 9)
}
```

### Out-of-fold confusion matrices

```{r xgboost-confusion-matrices}
prediction_files <- list.files(
  PREDICTIONS_DIR,
  pattern = sprintf("__%s__%s__xgboost\\.parquet$", SPACE, WINDOW),
  full.names = TRUE
)
if (length(prediction_files) == 0) {
  cat("No matching out-of-fold prediction files found.\n")
} else {
  confusion_data <- bind_rows(lapply(prediction_files, function(path) {
    target <- sub(sprintf("__%s__%s__xgboost\\.parquet$", SPACE, WINDOW), "", basename(path))
    as.data.frame(arrow::read_parquet(path)) %>%
      count(true_label, predicted_label, name = "n") %>%
      group_by(true_label) %>% mutate(row_fraction = n / sum(n)) %>% ungroup() %>%
      mutate(target = display_target(target))
  }))
  confusion_plot <- ggplot(confusion_data, aes(predicted_label, true_label, fill = row_fraction)) +
    geom_tile(colour = "white") +
    geom_text(aes(label = n), size = 3) +
    scale_fill_viridis_c(labels = function(x) paste0(round(100 * x), "%"), name = "Row fraction") +
    facet_wrap(~ target, scales = "free") +
    labs(title = "Out-of-fold confusion matrices", subtitle = "Rows are true labels; colors are normalized within true label",
         x = "Predicted label", y = "True label") + theme_report +
    theme(axis.text.x = element_text(angle = 45, hjust = 1))
  save_panel(confusion_plot, "10_xgboost_confusion_matrices", width = 14, height = 10)
}
```

## Reproducibility and interpretation notes

- PC scores use `L2 row normalization -> StandardScaler -> PCA` on concatenated clinician, imaging, and pathology means.
- PC association FDR is corrected within `(space, window, clinical family)` across all screened PCs and variables. Check `pc_join_coverage.csv` before interpreting a sparsely matched family.
- XGBoost models use nested stratified CV with blockwise PCA fit inside training partitions. Reported summaries and confusion matrices use outer-fold predictions only.
- The report writes each panel as PNG and PDF to `SEMANTIC_SEARCH_FIGURES_OUT`, or by default to `CLINICAL_FIGURES_OUT/semantic_search/<space>_<window>`.

Render, for example:

```{r render-command, eval=FALSE}
rmarkdown::render(
  "semantic_search/notebooks/05_figures.Rmd",
  params = list(space = "concat", window = "alltime", max_pcs = 20)
)
```
