---
title: "Generate code lookups (R tier)"
output: html_document
---

```{r setup, include=FALSE}
find_repo_root <- function() {
  d <- normalizePath(getwd(), mustWork = TRUE)
  while (nchar(d) > 1) {
    if (file.exists(file.path(d, "config.py")) &&
        dir.exists(file.path(d, "R"))) return(d)
    parent <- dirname(d)
    if (parent == d) break
    d <- parent
  }
  stop("Could not find repo root from ", getwd())
}

REPO_ROOT <- find_repo_root()
knitr::opts_knit$set(root.dir = REPO_ROOT)
```

Runs the two `pipelines/preprocessing/generate_*.R` scripts that build the phecode lookup tables
`figures.prep.figure2` labels its panels from. This is the **only** R dependency in the otherwise
Python figure-data prep tier, split out here so `4_figures/02` runs under a plain Python kernel with no
`Rscript` on `PATH`.

Like the other R scripts, both generators `source("R/config.R")` with a relative path, so this
document sets the R working directory to the repo root (via `knitr::opts_knit$set(root.dir = ...)`,
which — unlike a bare `setwd()` in a chunk — persists across all chunks) before sourcing anything.

Writes to `CODE_PATH` (`$DATA_PATH/code_data/`):

| Output | Used by |
|---|---|
| `icd10_to_phecode_mapping.csv` | `figures.prep.figure2` (cross-scheme event dedup + phecode labels), `pipelines.preprocessing.generate_embedding_prediction_datasets` |
| `icd10_unmapped_codes.csv` | diagnostic only — codes no phecode covers |
| `phecode_descriptions.csv` | `figures.prep.figure2` (phecode label fallback) |

**These outputs are static reference data.** They change only when the Phecode package tables or
the cohort's set of unique ICD-10 codes change, so this document is a **one-time bootstrap** —
re-run it after a cohort rebuild (`1_data/01`) or a Phecode package upgrade, not before
every `4_figures/02`.

- **Prep tier**: [02_figure_data.ipynb](02_figure_data.ipynb) (Python kernel) — run after this.
- **Render tier**: [03_render_figures.Rmd](03_render_figures.Rmd) (R Markdown).

## Prerequisites

Beyond the figure-rendering packages (`Rscript R/install_packages.R`), these generators need
`arrow` and the Phecode data package, which is GitHub-only:

```
install.packages(c("data.table", "arrow", "devtools"))
devtools::install_github("vcastro/Phecode")
```

`generate_icd10_to_phecode_mapping.R` also reads `SURV_PATH/timestamped_icd_info.parquet`, so
step `1_data/01` must have run.

```{r preflight}
source(file.path(REPO_ROOT, "R", "config.R"))

# Set to TRUE to re-run the generators even when their outputs already exist.
FORCE_REGENERATE <- FALSE

GENERATORS <- list(
  list(script = "generate_icd10_to_phecode_mapping.R", output = "icd10_to_phecode_mapping.csv"),
  list(script = "generate_phecode_descriptions.R",     output = "phecode_descriptions.csv")
)

cat("repo root:   ", REPO_ROOT, "\n", sep = "")
cat("CODE_PATH: ", CODE_PATH, "\n", sep = "")
cat("R version: ", R.version.string, "\n\n", sep = "")

for (g in GENERATORS) {
  path <- file.path(CODE_PATH, g$output)
  if (file.exists(path)) {
    cat(sprintf("  [ok] %-32s %s bytes\n", g$output, format(file.size(path), big.mark = ",")))
  } else {
    cat(sprintf("  [--] %-32s missing\n", g$output))
  }
}

missing_pkgs <- Filter(function(p) !requireNamespace(p, quietly = TRUE),
                       c("Phecode", "data.table", "arrow"))
if (length(missing_pkgs) > 0) {
  cat("\nMissing R packages: ", paste(missing_pkgs, collapse = ", "),
      "\n  See the prerequisites above.\n", sep = "")
}
```

```{r generate}
icd_parquet <- file.path(SURV_PATH, "timestamped_icd_info.parquet")
if (!file.exists(icd_parquet)) {
  stop(icd_parquet, " not found - run notebook 1_data/01 before generating the ICD->phecode mapping.")
}

# Each generator is sys.source()-ed in its own environment so their identically-named helpers
# (normalize_phecode(), find_col(), ...) don't leak between them, matching 4_figures/03's convention.
for (g in GENERATORS) {
  script_path <- file.path(REPO_ROOT, "pipelines", "preprocessing", g$script)
  if (!file.exists(script_path)) stop("Code lookup generator not found: ", script_path)

  out_path <- file.path(CODE_PATH, g$output)
  if (file.exists(out_path) && !FORCE_REGENERATE) {
    cat("\n=== ", g$script, " - SKIPPED (", g$output,
        " exists; set FORCE_REGENERATE <- TRUE to rebuild) ===\n", sep = "")
    next
  }

  cat("\n=== ", g$script, " ===\n", sep = "")
  env <- new.env(parent = globalenv())
  sys.source(script_path, envir = env)
}

cat("\nDone. Code lookups written to ", CODE_PATH, "\n", sep = "")
cat("Next: 02_figure_data.ipynb\n")
```

```{r verify}
# Verify what 4_figures/02 will actually pick up.
mapping_path <- file.path(CODE_PATH, "icd10_to_phecode_mapping.csv")
desc_path <- file.path(CODE_PATH, "phecode_descriptions.csv")

if (file.exists(mapping_path)) {
  mapping <- data.table::fread(mapping_path, colClasses = "character")
  cat(sprintf("icd10_to_phecode_mapping.csv: %s rows, %s ICD-10 codes, %s phecodes\n",
              format(nrow(mapping), big.mark = ","),
              format(length(unique(mapping$icd10_code)), big.mark = ","),
              format(length(unique(mapping$phecode)), big.mark = ",")))
  print(utils::head(mapping, 5))
} else {
  cat("icd10_to_phecode_mapping.csv MISSING - figure2 will disable cross-scheme event",
      "dedup and fall back to raw codes for phecode labels.\n")
}

cat("\n")
if (file.exists(desc_path)) {
  desc <- data.table::fread(desc_path, colClasses = "character")
  cat(sprintf("phecode_descriptions.csv: %s rows\n", format(nrow(desc), big.mark = ",")))
  print(utils::head(desc, 5))
} else {
  cat("phecode_descriptions.csv MISSING - figure2 phecode panels will show bare codes.\n")
}
```

## Running without R

`4_figures/02` does not hard-fail on missing lookups — `figures.prep.figure2` falls back to raw codes and
logs the miss counts (`_report_lookup_misses()`) at the end of its run. What you lose:

- **Cross-scheme event dedup is disabled** — `icd3_post`/`icd4_post`/`phecode_post` events that
  denote the same phenotype are no longer collapsed.
- **Phecode panels show bare codes** (`250.2`) instead of descriptions (`Type 2 diabetes`).

That is fine for a diagnostic run and **not** fine for manuscript figures. If you need final
figures on a machine without R, generate these CSVs elsewhere and copy them into `CODE_PATH`.

Render with:

```
Rscript -e 'rmarkdown::render("notebooks/4_figures/01_code_lookups.Rmd")'
```
