---
title: "Render manuscript figures (R tier)"
output: html_document
---

```{r setup, include=FALSE}
find_repo_root <- function() {
  d <- normalizePath(getwd(), mustWork = TRUE)
  while (nchar(d) > 1) {
    if (file.exists(file.path(d, "config.py")) &&
        dir.exists(file.path(d, "R"))) return(d)
    parent <- dirname(d)
    if (parent == d) break
    d <- parent
  }
  stop("Could not find repo root from ", getwd())
}

REPO_ROOT <- find_repo_root()
knitr::opts_knit$set(root.dir = REPO_ROOT)
```

Runs the `R/plot_figure_*.R` scripts in this notebook's **R kernel**. Each
script reads the CSVs written by the Python prep tier and builds every panel as a standalone
ggplot object, saving each one individually (both PNG and PDF) — there is no composed
multi-panel figure.

Metric-aware figures render twice: once using mean time-dependent AUC(t), the primary metric,
and once using Harrell's C-index. Metric-invariant figures render once.

A script belongs in `METRIC_SCRIPTS` if **any** of its panels depends on the active metric —
including only through the event-exclusion set, which is judged on that metric (this is why
figure 1 is metric-aware: `fig1b` recounts endpoints after the exclusion). Such panels are saved
metric-tagged, so listing the script as metric-invariant silently renders just the AUC variant and
never writes the C-index file.

- **Code lookups (R)**: [01_code_lookups.Rmd](01_code_lookups.Rmd) — one-time bootstrap.
- **Prep tier**: [02_figure_data.ipynb](02_figure_data.ipynb) (Python kernel) — run first.
- **One-time R package bootstrap** (any host with R):

  ```
  Rscript R/install_packages.R
  ```

Each plot script is `sys.source()`-ed in its own environment so per-script ggplot objects
(`p2a`, `p3b`, …) don't leak between cells. R scripts always assume the working directory
is the repo root (they `source("R/figure_utils.R")` with a relative path), so this document sets the
R working directory to the repo root (via `knitr::opts_knit$set(root.dir = ...)`, which — unlike a
bare `setwd()` in a chunk — persists across all chunks) before sourcing anything.

```{r find-root}
R_DIR <- file.path(REPO_ROOT, "R")

cat("repo root:   ", REPO_ROOT, "\n", sep = "")
cat("R scripts: ", R_DIR, "\n", sep = "")
cat("R version: ", R.version.string, "\n", sep = "")
```

```{r render-plots}
METRIC_INVARIANT_SCRIPTS <- c(
  "plot_figure_0.R",
  "plot_figure_4.R",
  "plot_figure_4_supp.R",
  "plot_figure_5.R"
)

# plot_figure_1.R belongs here, not in METRIC_INVARIANT_SCRIPTS: fig1b's endpoint
# counts are recounted after the event exclusion, which is judged on the active
# metric, so the panel differs between metrics and is saved metric-tagged
# (fig1b_cindex / fig1b_auc). Rendered as metric-invariant it ran once with
# MANUSCRIPT_METRIC unset -- which figure_utils.R defaults to "auc" -- and
# fig1b_cindex was never written at all. Figure 1's other panels have no event
# dimension and are simply rewritten identically by the second pass.
METRIC_SCRIPTS <- c(
  "plot_figure_1.R",
  "plot_figure_2.R",
  "plot_figure_2_supp.R",
  "plot_figure_2_supp_events.R",
  "plot_figure_2_supp_anchor.R",
  "plot_figure_3.R"
)

run_plot <- function(script, metric = NULL) {
  if (is.null(metric)) Sys.unsetenv("MANUSCRIPT_METRIC") else Sys.setenv(MANUSCRIPT_METRIC = metric)
  cat("\n=== ", script, if (is.null(metric)) "" else paste0(" [", metric, "]"), " ===\n", sep = "")
  env <- new.env(parent = globalenv())
  sys.source(file.path(R_DIR, script), envir = env)
}

for (s in METRIC_INVARIANT_SCRIPTS) run_plot(s)
for (metric in c("auc", "cindex")) {
  for (s in METRIC_SCRIPTS) run_plot(s, metric)
}

Sys.unsetenv("MANUSCRIPT_METRIC")

cat("\nDone.\n")
cat("Panel PNGs: $CLINICAL_FIGURES_OUT/png/<figureN>/ (or /data/.../manuscript_figures/png/<figureN>/)\n")
cat("Panel PDFs: $CLINICAL_FIGURES_OUT/pdf/<figureN>/ (or /data/.../manuscript_figures/pdf/<figureN>/)\n")
```

Render with:

```
Rscript -e 'rmarkdown::render("notebooks/4_figures/03_render_figures.Rmd")'
```
