---
title: "Render manuscript figures (R tier)"
output: html_document
---

```{r setup, include=FALSE}
find_repo_root <- function() {
  d <- normalizePath(getwd(), mustWork = TRUE)
  while (nchar(d) > 1) {
    if (file.exists(file.path(d, "config.py")) &&
        dir.exists(file.path(d, "R"))) return(d)
    parent <- dirname(d)
    if (parent == d) break
    d <- parent
  }
  stop("Could not find repo root from ", getwd())
}

REPO_ROOT <- find_repo_root()
knitr::opts_knit$set(root.dir = REPO_ROOT)
```

Runs the `R/plot_figure_*.R` scripts in this notebook's **R kernel**. Each
script reads the CSVs written by the Python prep tier and saves individual panels
plus compiled Figures 1–4 in PNG and PDF. Compiled figures carry lowercase panel
labels (a, b, c, …), with one label per named panel, including panels with internal
facets or subplots. Missing panels prevent a compiled figure from being published.

All performance plots use Harrell's C-index. AUC variants and the former Figure 5a
propensity ROC panel are no longer rendered. Old AUC/ROC image files in the output
folders are removed when the rendering utilities are loaded.

- **Code lookups (R)**: [01_code_lookups.Rmd](01_code_lookups.Rmd) — one-time bootstrap.
- **Prep tier**: [02_figure_data.ipynb](02_figure_data.ipynb) (Python kernel) — run first.
- **One-time R package bootstrap** (any host with R):

  ```
  Rscript R/install_packages.R
  ```

Each plot script is `sys.source()`-ed in its own environment so per-script ggplot objects
(`p2a`, `p3b`, …) don't leak between cells. R scripts always assume the working directory
is the repo root (they `source("R/figure_utils.R")` with a relative path), so this document sets the
R working directory to the repo root (via `knitr::opts_knit$set(root.dir = ...)`, which — unlike a
bare `setwd()` in a chunk — persists across all chunks) before sourcing anything.

```{r find-root}
R_DIR <- file.path(REPO_ROOT, "R")

cat("repo root:   ", REPO_ROOT, "\n", sep = "")
cat("R scripts: ", R_DIR, "\n", sep = "")
cat("R version: ", R.version.string, "\n", sep = "")
```

```{r render-plots}
SCRIPTS <- c(
  "plot_figure_0.R",
  "plot_figure_1.R",
  "plot_figure_2.R",
  "plot_figure_2_supp.R",
  "plot_figure_2_supp_events.R",
  "plot_figure_2_supp_anchor.R",
  "plot_figure_2_supp_family.R",
  "plot_figure_3.R",
  "plot_figure_4.R",
  "plot_figure_4_supp.R",
  "plot_figure_5.R"
)

for (script in SCRIPTS) {
  cat("\n=== ", script, " [cindex] ===\n", sep = "")
  env <- new.env(parent = globalenv())
  sys.source(file.path(R_DIR, script), envir = env)
}

cat("\nDone.\n")
cat("Panel PNGs: $CLINICAL_FIGURES_OUT/png/<figureN>/ (or /data/.../manuscript_figures/png/<figureN>/)\n")
cat("Compiled figures: figureN_cindex.png/.pdf alongside each figure’s panels\n")
cat("Panel PDFs: $CLINICAL_FIGURES_OUT/pdf/<figureN>/ (or /data/.../manuscript_figures/pdf/<figureN>/)\n")
```

Render with:

```
Rscript -e 'rmarkdown::render("notebooks/4_figures/03_render_figures.Rmd")'
```
