---
title: "Render manuscript figures (R tier)"
output: html_document
---

```{r setup, include=FALSE}
find_repo_root <- function() {
  d <- normalizePath(getwd(), mustWork = TRUE)
  while (nchar(d) > 1) {
    if (file.exists(file.path(d, "config.py")) &&
        dir.exists(file.path(d, "R"))) return(d)
    parent <- dirname(d)
    if (parent == d) break
    d <- parent
  }
  stop("Could not find repo root from ", getwd())
}

REPO_ROOT <- find_repo_root()
knitr::opts_knit$set(root.dir = REPO_ROOT)
```

Runs the `R/plot_figure_*.R` scripts in this notebook's **R kernel**. Each
script reads the CSVs written by the Python prep tier and saves individual panels
plus compiled Figures 1–4 in PNG and PDF. Compiled figures carry lowercase panel
labels (a, b, c, …), with one label per named panel, including panels with internal
facets or subplots. Missing panels are skipped with their reasons logged. Available
panels retain their original letters, and the compiled figure lists unavailable
panels in its caption. A figure with no available panels is skipped without
stopping the remaining figures.

All performance plots use Harrell's C-index. AUC variants and the former Figure 5a
propensity ROC panel are no longer rendered. Old AUC/ROC image files in the output
folders are removed when the rendering utilities are loaded.

The main figures contain exactly these panels, in letter order:

- **Figure 1 (a–d):** cancer-type pie, stage barplot, cohort availability flowchart,
  outcome endpoints.
- **Figure 2 (a–d):** base-versus-text C-index scatter, delta violin, stage KM,
  text-risk quartile KM.
- **Figure 3 (a–c):** modality rank, significant endpoint barplot, joint Cox violins.
- **Figure 4 (a–d):** risk-score heatmap, KM by risk trajectory, risk-dynamics
  composition by stage, stage I–II rising-risk versus stage IV falling-risk KM.

Within-versus-pan and top-event plots (including the top-event supplement) are
retired. Superseded standalone panel files are removed when each figure rerenders.

Main figures are rendered at 180-mm width with final-size type, 600-dpi PNGs, and
vector PDFs. Each compiled figure also gets a `_captioned` PNG/PDF review copy;
standalone legends are written to `$CLINICAL_FIGURES_OUT/captions/`. Base wording
is versioned in `figures/manuscript_captions.json`; current sample sizes and model
statistics are added at rendering time. The clean artwork remains available for
journals requiring captions separately.

- **Code lookups (R)**: [01_code_lookups.Rmd](01_code_lookups.Rmd) — one-time bootstrap.
- **Prep tier**: [02_figure_data.ipynb](02_figure_data.ipynb) (Python kernel) — run first.
- **One-time R package bootstrap** (any host with R):

  ```
  Rscript R/install_packages.R
  ```

Each plot script is `sys.source()`-ed in its own environment so per-script ggplot objects
(`p2a`, `p3b`, …) don't leak between cells. R scripts always assume the working directory
is the repo root (they `source("R/figure_utils.R")` with a relative path), so this document sets the
R working directory to the repo root (via `knitr::opts_knit$set(root.dir = ...)`, which — unlike a
bare `setwd()` in a chunk — persists across all chunks) before sourcing anything.

```{r find-root}
R_DIR <- file.path(REPO_ROOT, "R")

cat("repo root:   ", REPO_ROOT, "\n", sep = "")
cat("R scripts: ", R_DIR, "\n", sep = "")
cat("R version: ", R.version.string, "\n", sep = "")
```

```{r render-plots}
# Reset both legacy controls in an existing R session before sourcing scripts.
# This also keeps older plot-script copies from inheriting an AUC selection.
Sys.setenv(MANUSCRIPT_METRIC = "cindex", MANUSCRIPT_METRICS = "cindex")
cat("C-index-only renderer: ", file.path(R_DIR, "..", "notebooks", "4_figures",
                                      "03_render_figures.Rmd"), "\n", sep = "")

SCRIPTS <- c(
  "plot_figure_0.R",
  "plot_figure_1.R",
  "plot_figure_2.R",
  "plot_figure_2_supp.R",
  "plot_figure_2_supp_anchor.R",
  "plot_figure_2_supp_family.R",
  "plot_figure_3.R",
  "plot_figure_4.R",
  "plot_figure_4_supp.R",
  "plot_figure_5.R"
)

for (script in SCRIPTS) {
  cat("\n=== ", script, " [cindex] ===\n", sep = "")
  env <- new.env(parent = globalenv())
  sys.source(file.path(R_DIR, script), envir = env)
}

cat("\nDone.\n")
cat("Panel PNGs: $CLINICAL_FIGURES_OUT/png/<figureN>/ (or /data/.../manuscript_figures/png/<figureN>/)\n")
cat("Compiled figures: figureN_cindex.png/.pdf alongside each figure’s panels\n")
cat("Panel PDFs: $CLINICAL_FIGURES_OUT/pdf/<figureN>/ (or /data/.../manuscript_figures/pdf/<figureN>/)\n")
```

Render with:

```
Rscript -e 'rmarkdown::render("notebooks/4_figures/03_render_figures.Rmd")'
```
