---
title: "Biomarker hits: Kaplan-Meier curves"
output:
  html_document:
    toc: true
    toc_float: true
---

This report renders every FDR-significant marker-by-ICI hit twice: an unweighted
Kaplan-Meier estimate and a stabilized, truncated ATE-weighted estimate on the
same common-support-trimmed population used by the biomarker screen. In addition
to the PNGs, it writes compiled, long-format CSVs for every curve and at-risk
table, plus an export manifest.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

find_repo_root <- function(start = getwd()) {
  candidate <- normalizePath(start, mustWork = TRUE)
  repeat {
    if (file.exists(file.path(candidate, "config.py")) &&
        dir.exists(file.path(candidate, "pipelines"))) return(candidate)
    parent <- dirname(candidate)
    if (identical(parent, candidate)) stop("Could not find repository root from ", start)
    candidate <- parent
  }
}

REPO_ROOT <- find_repo_root()
knitr::opts_knit$set(root.dir = REPO_ROOT)
setwd(REPO_ROOT)

suppressPackageStartupMessages({
  library(arrow)
  library(dplyr)
  library(ggplot2)
  library(patchwork)
  library(readr)
  library(survival)
  library(tidyr)
})
source("R/config.R")

BIOMARKER_PATH <- file.path(DATA_PATH, "biomarker_analysis")
COMPILED_DIR <- file.path(BIOMARKER_PATH, "compiled_results")
HITS_FILE <- file.path(COMPILED_DIR, "track2_all_significant_hits.csv")
KM_OUT_DIR <- Sys.getenv(
  "KM_OUT_DIR",
  unset = file.path(FIGURE_OUT_DIR, "figure5", "KM_curves")
)

COMMON_SUPPORT_PCT <- c(0.5, 99.5)
IPTW_TRUNC_PCT <- c(1, 99)
TIME_DIVISOR <- 30.44
MAX_FOLLOWUP <- 60
RISK_TIMES <- seq(0, MAX_FOLLOWUP, length.out = 6)
DPI <- 200
OVERWRITE <- tolower(Sys.getenv("KM_OVERWRITE", "true")) %in% c("1", "true", "yes")

GROUP_LEVELS <- c(
  "marker- / non-ICI", "marker+ / non-ICI",
  "marker- / ICI", "marker+ / ICI"
)
GROUP_COLORS <- c("marker-" = "#4C72B0", "marker+" = "#C44E52")
GROUP_LINES <- c("non-ICI" = "dashed", "ICI" = "solid")

dir.create(KM_OUT_DIR, recursive = TRUE, showWarnings = FALSE)
```

## Inputs and hit consolidation

One panel is generated per unique `(marker, cancer_type, cohort, ps_model)`.
When both weighted screens call the same hit, `sig_in` records both calls.

```{r inputs}
if (!file.exists(HITS_FILE)) {
  stop(HITS_FILE, " not found. Run stage 6 (compile_IPTW_results) first.")
}

hits <- tryCatch(readr::read_csv(HITS_FILE, show_col_types = FALSE),
                 error = function(e) tibble())

if (nrow(hits) == 0) {
  panels <- tibble()
} else {
  if (!"extreme_hr_flag" %in% names(hits)) hits$extreme_hr_flag <- FALSE
  panels <- hits %>%
    mutate(HR_markerxICI = as.numeric(HR_markerxICI),
           p_markerxICI = as.numeric(p_markerxICI)) %>%
    group_by(marker, cancer_type, cohort, ps_model) %>%
    summarise(
      sig_in = paste(sort(unique(weight_type)), collapse = "+"),
      min_p = suppressWarnings(min(p_markerxICI, na.rm = TRUE)),
      HR_first = first(HR_markerxICI),
      extreme_hr = any(extreme_hr_flag, na.rm = TRUE),
      .groups = "drop"
    ) %>%
    mutate(min_p = ifelse(is.infinite(min_p), NA_real_, min_p)) %>%
    arrange(min_p)
}

cat(nrow(panels), "unique hit panels\n")
```

## Population reconstruction

For cancer-specific hits, the propensity score is recalibrated within cancer
type by fitting treatment on the original propensity score. Both arms are then
restricted to their overlapping 0.5th--99.5th percentile support. Stabilized ATE
weights are truncated at their 1st and 99th percentiles. These constants match
`pipelines/biomarkers/run_IPTW_analysis.py`.

```{r helpers}
safe_name <- function(x) gsub("[^[:alnum:]_.-]", "_", as.character(x))

effective_sample_size <- function(w) {
  denom <- sum(w ^ 2)
  if (denom > 0) sum(w) ^ 2 / denom else 0
}

group_label <- function(marker_value, treated) {
  paste0(ifelse(marker_value == 1, "marker+", "marker-"), " / ",
         ifelse(treated == 1, "ICI", "non-ICI"))
}

load_hit_population <- function(marker, cancer_type, cohort, ps_model) {
  fp <- file.path(BIOMARKER_PATH, paste0("IPTW_df_", cohort, "_", ps_model, ".parquet"))
  if (!file.exists(fp)) stop(fp, " not found -- run stage 4 (generate_IPTW_df).")
  df <- as.data.frame(arrow::read_parquet(fp))
  if (!marker %in% names(df)) stop("marker '", marker, "' is absent from ", basename(fp))

  n_full <- nrow(df)
  if (cancer_type != "pan_cancer") {
    ct_col <- paste0("CANCER_TYPE_", cancer_type)
    if (!ct_col %in% names(df)) stop(ct_col, " is absent from ", basename(fp))
    df <- df[!is.na(df[[ct_col]]) & as.logical(df[[ct_col]]), , drop = FALSE]
  }
  n_cancer_type <- nrow(df)

  required <- c(marker, "PX_on_ICI", "tt_death", "death", "ICI_prediction")
  for (nm in required) df[[nm]] <- suppressWarnings(as.numeric(df[[nm]]))
  keep <- Reduce(`&`, lapply(df[required], is.finite)) & df$tt_death > 0
  df <- df[keep, , drop = FALSE]
  df$marker_value <- as.integer(df[[marker]] > 0)
  n_model_frame <- nrow(df)
  if (nrow(df) == 0 || length(unique(df$PX_on_ICI)) < 2) {
    stop("no rows, or only one treatment arm, before trimming")
  }

  recalibrated <- FALSE
  if (cancer_type != "pan_cancer" && nrow(df) > 10) {
    ps_fit <- glm(PX_on_ICI ~ ICI_prediction, data = df, family = binomial())
    df$ICI_prediction <- predict(ps_fit, newdata = df, type = "response")
    recalibrated <- TRUE
  }

  ps_raw <- pmin(pmax(df$ICI_prediction, 1e-6), 1 - 1e-6)
  ps_t <- ps_raw[df$PX_on_ICI == 1]
  ps_c <- ps_raw[df$PX_on_ICI == 0]
  lower <- max(quantile(ps_t, COMMON_SUPPORT_PCT[1] / 100, names = FALSE, type = 7),
               quantile(ps_c, COMMON_SUPPORT_PCT[1] / 100, names = FALSE, type = 7))
  upper <- min(quantile(ps_t, COMMON_SUPPORT_PCT[2] / 100, names = FALSE, type = 7),
               quantile(ps_c, COMMON_SUPPORT_PCT[2] / 100, names = FALSE, type = 7))
  if (!is.finite(lower) || !is.finite(upper) || lower >= upper) {
    stop("no propensity overlap; cannot reproduce the analysis population")
  }

  df <- df[ps_raw >= lower & ps_raw <= upper, , drop = FALSE]
  if (nrow(df) == 0 || length(unique(df$PX_on_ICI)) < 2) {
    stop("no rows, or only one arm, after common-support trimming")
  }

  ps <- pmin(pmax(df$ICI_prediction, 1e-6), 1 - 1e-6)
  p_treated <- mean(df$PX_on_ICI)
  w <- ifelse(df$PX_on_ICI == 1, p_treated / ps, (1 - p_treated) / (1 - ps))
  limits <- quantile(w, IPTW_TRUNC_PCT / 100, names = FALSE, type = 7)
  df$IPTW_ATE <- pmin(pmax(w, limits[1]), limits[2])
  df$time_months <- df$tt_death / TIME_DIVISOR
  df$group <- factor(group_label(df$marker_value, df$PX_on_ICI), levels = GROUP_LEVELS)

  counts <- df %>% count(group, name = "n") %>%
    complete(group = factor(GROUP_LEVELS, levels = GROUP_LEVELS), fill = list(n = 0))
  if (any(counts$n == 0)) {
    stop("no patients in ", paste(as.character(counts$group[counts$n == 0]), collapse = ", "),
         " after common-support trimming")
  }

  list(
    data = df,
    info = list(n_full = n_full, n_cancer_type = n_cancer_type,
                n_model_frame = n_model_frame, n_analysis = nrow(df),
                n_trimmed_out = n_model_frame - nrow(df), recalibrated = recalibrated,
                ps_lower = lower, ps_upper = upper,
                ess_treated = effective_sample_size(df$IPTW_ATE[df$PX_on_ICI == 1]),
                ess_control = effective_sample_size(df$IPTW_ATE[df$PX_on_ICI == 0]))
  )
}

fit_one_curve <- function(df, weighted, hit_keys) {
  bind_rows(lapply(GROUP_LEVELS, function(g) {
    sub <- df[df$group == g, , drop = FALSE]
    weights <- if (weighted) sub$IPTW_ATE else rep(1, nrow(sub))
    fit <- survfit(Surv(time_months, death) ~ 1, data = sub, weights = weights,
                   conf.type = "log")
    s <- summary(fit, censored = FALSE)
    tibble(
      !!!hit_keys,
      weighting = ifelse(weighted, "IPTW_ATE", "unweighted"),
      group = g,
      time = c(0, s$time), survival = c(1, s$surv),
      n_risk = c(nrow(sub), s$n.risk), n_event = c(0, s$n.event),
      std_error = c(0, s$std.err), lower = c(1, s$lower), upper = c(1, s$upper)
    )
  }))
}

weighted_logrank <- function(df, weighted = FALSE) {
  x <- df[df$PX_on_ICI == 1, , drop = FALSE]
  if (length(unique(x$marker_value)) < 2) return(NA_real_)
  w <- if (weighted) x$IPTW_ATE else rep(1, nrow(x))
  event_times <- sort(unique(x$time_months[x$death == 1]))
  score <- 0
  variance <- 0
  for (tt in event_times) {
    at_risk <- x$time_months >= tt
    at_event <- x$time_months == tt & x$death == 1
    y1 <- sum(w[at_risk & x$marker_value == 1])
    y0 <- sum(w[at_risk & x$marker_value == 0])
    d1 <- sum(w[at_event & x$marker_value == 1])
    d <- sum(w[at_event])
    y <- y1 + y0
    if (y > 0) score <- score + d1 - d * y1 / y
    if (y > 1) variance <- variance + y1 * y0 * d * (y - d) / (y ^ 2 * (y - 1))
  }
  if (variance <= 0) NA_real_ else pchisq(score ^ 2 / variance, df = 1, lower.tail = FALSE)
}

make_at_risk <- function(df, hit_keys) {
  bind_rows(lapply(GROUP_LEVELS, function(g) {
    sub <- df[df$group == g, , drop = FALSE]
    bind_rows(lapply(RISK_TIMES, function(tt) {
      at_risk <- sub$time_months >= tt
      tibble(!!!hit_keys, group = g, time = tt,
             n_at_risk = sum(at_risk),
             weighted_at_risk = sum(sub$IPTW_ATE[at_risk]))
    }))
  }))
}

render_hit <- function(df, curves, risk, row, info, p_unweighted, p_weighted, output_file) {
  curves <- curves %>%
    mutate(marker_status = ifelse(grepl("marker\\+", group), "marker+", "marker-"),
           ici_status = ifelse(grepl("non-ICI", group), "non-ICI", "ICI"))

  group_stats <- risk %>% filter(time == 0) %>% select(group, n_at_risk)
  panel_plot <- function(weighting, p_value) {
    dd <- curves %>% filter(.data$weighting == weighting) %>%
      left_join(group_stats, by = "group")
    labels <- setNames(vapply(GROUP_LEVELS, function(g) {
      sub <- df[df$group == g, , drop = FALSE]
      if (weighting == "IPTW_ATE") {
        sprintf("%s (n=%d, ESS=%.0f, d=%d)", g, nrow(sub),
                effective_sample_size(sub$IPTW_ATE), sum(sub$death))
      } else {
        sprintf("%s (n=%d, d=%d)", g, nrow(sub), sum(sub$death))
      }
    }, character(1)), GROUP_LEVELS)
    group_colors <- setNames(ifelse(grepl("marker\\+", GROUP_LEVELS),
                                    GROUP_COLORS[["marker+"]], GROUP_COLORS[["marker-"]]),
                             GROUP_LEVELS)
    group_lines <- setNames(ifelse(grepl("non-ICI", GROUP_LEVELS), "dashed", "solid"),
                            GROUP_LEVELS)
    ggplot(dd, aes(time, survival, color = group, linetype = group,
                   group = group)) +
      geom_step(linewidth = 0.8) +
      scale_color_manual(values = group_colors, labels = labels) +
      scale_linetype_manual(values = group_lines, labels = labels) +
      coord_cartesian(xlim = c(0, MAX_FOLLOWUP), ylim = c(0, 1.02)) +
      labs(x = "Time (months)", y = "Survival probability",
           title = ifelse(weighting == "IPTW_ATE", "IPTW-weighted (ATE)", "Unweighted"),
           subtitle = paste0("marker+/- within ICI: log-rank p=",
                             ifelse(is.na(p_value), "n/a", format.pval(p_value, digits = 3))),
           color = NULL, linetype = NULL) +
      guides(linetype = "none",
             color = guide_legend(override.aes = list(linetype = unname(group_lines)))) +
      theme_bw(base_size = 10) + theme(legend.position = "bottom")
  }

  risk_plot <- risk %>%
    mutate(group = factor(group, levels = rev(GROUP_LEVELS))) %>%
    ggplot(aes(time, group, label = n_at_risk)) +
    geom_text(size = 3) +
    scale_x_continuous(limits = c(0, MAX_FOLLOWUP), breaks = RISK_TIMES) +
    labs(x = "Time (months)", y = NULL, title = "Number at risk") +
    theme_minimal(base_size = 9) +
    theme(panel.grid = element_blank(), axis.text.y = element_text(hjust = 0))

  title_bits <- c(
    paste0(row$marker, " — ", row$cancer_type),
    paste0(row$cohort, " / ", row$ps_model,
           " | HR(marker×ICI)=", ifelse(is.finite(row$HR_first), sprintf("%.2f", row$HR_first), "n/a"),
           " | p=", ifelse(is.finite(row$min_p), format.pval(row$min_p, digits = 2), "n/a"),
           " | FDR-sig in: ", row$sig_in),
    paste0("n=", info$n_analysis, " after common-support trim (-", info$n_trimmed_out,
           " of ", info$n_model_frame, ")")
  )
  composed <- (panel_plot("unweighted", p_unweighted) |
                 panel_plot("IPTW_ATE", p_weighted)) /
    risk_plot +
    plot_layout(heights = c(4, 1.35)) +
    plot_annotation(title = paste(title_bits, collapse = "\n"))
  ggsave(output_file, composed, width = 13.5, height = 7.2, dpi = DPI, bg = "white")
}
```

## Export

The long-format CSVs are regenerated atomically at the end of each successful
report run:

- `KM_compiled_curves.csv`: survival estimates and uncertainty at every event time.
- `KM_compiled_at_risk.csv`: unweighted and weighted at-risk totals at fixed plot times.
- `KM_export_manifest.csv`: one row per requested hit, including skips, group support,
  effective sample size, propensity support, and within-ICI log-rank p-values.

```{r export}
curve_parts <- list()
risk_parts <- list()
manifest_parts <- list()

for (i in seq_len(nrow(panels))) {
  row <- panels[i, , drop = FALSE]
  keys <- as.list(row[c("marker", "cancer_type", "cohort", "ps_model")])
  spec <- paste(row$cohort, row$ps_model, sep = "_")
  spec_dir <- file.path(KM_OUT_DIR, spec)
  output_file <- file.path(spec_dir, paste0("KM_", safe_name(row$marker), "_",
                                            safe_name(row$cancer_type), ".png"))
  base_record <- row %>% transmute(
    marker, cancer_type, cohort, ps_model, sig_in,
    HR_markerxICI = HR_first, p_markerxICI = min_p,
    extreme_hr_flag = extreme_hr
  )

  result <- tryCatch(load_hit_population(row$marker, row$cancer_type,
                                         row$cohort, row$ps_model),
                     error = function(e) e)
  if (inherits(result, "error")) {
    manifest_parts[[length(manifest_parts) + 1]] <- base_record %>%
      mutate(status = "skipped", reason = conditionMessage(result), file = NA_character_)
    message(sprintf("[%d/%d] SKIP %s/%s/%s: %s", i, nrow(panels), spec,
                    row$marker, row$cancer_type, conditionMessage(result)))
    next
  }

  df <- result$data
  info <- result$info
  curves <- bind_rows(fit_one_curve(df, FALSE, keys), fit_one_curve(df, TRUE, keys))
  risk <- make_at_risk(df, keys)
  p_unweighted <- weighted_logrank(df, FALSE)
  p_weighted <- weighted_logrank(df, TRUE)

  status <- "written"
  if (!OVERWRITE && file.exists(output_file)) {
    status <- "existing"
  } else {
    dir.create(spec_dir, recursive = TRUE, showWarnings = FALSE)
    render_hit(df, curves, risk, row, info, p_unweighted, p_weighted, output_file)
  }

  group_summary <- df %>% group_by(group) %>%
    summarise(n = n(), deaths = sum(death), .groups = "drop")
  get_group <- function(g, field) group_summary[[field]][match(g, group_summary$group)]

  manifest_parts[[length(manifest_parts) + 1]] <- base_record %>% mutate(
    status = status, reason = NA_character_, file = file.path(spec, basename(output_file)),
    n_analysis = info$n_analysis, n_trimmed_out = info$n_trimmed_out,
    propensity_recalibrated = info$recalibrated,
    propensity_support_lower = info$ps_lower, propensity_support_upper = info$ps_upper,
    ess_treated = info$ess_treated, ess_control = info$ess_control,
    logrank_p_unweighted = p_unweighted, logrank_p_IPTW_ATE = p_weighted,
    n_marker_neg_nonICI = get_group("marker- / non-ICI", "n"),
    n_marker_pos_nonICI = get_group("marker+ / non-ICI", "n"),
    n_marker_neg_ICI = get_group("marker- / ICI", "n"),
    n_marker_pos_ICI = get_group("marker+ / ICI", "n"),
    d_marker_neg_nonICI = get_group("marker- / non-ICI", "deaths"),
    d_marker_pos_nonICI = get_group("marker+ / non-ICI", "deaths"),
    d_marker_neg_ICI = get_group("marker- / ICI", "deaths"),
    d_marker_pos_ICI = get_group("marker+ / ICI", "deaths")
  )
  curve_parts[[length(curve_parts) + 1]] <- curves
  risk_parts[[length(risk_parts) + 1]] <- risk
  message(sprintf("[%d/%d] %s %s/%s", i, nrow(panels), status, spec, basename(output_file)))
}

curves_compiled <- if (length(curve_parts)) bind_rows(curve_parts) else tibble(
  marker = character(), cancer_type = character(), cohort = character(), ps_model = character(),
  weighting = character(), group = character(), time = double(), survival = double(),
  n_risk = double(), n_event = double(), std_error = double(), lower = double(), upper = double()
)
at_risk_compiled <- if (length(risk_parts)) bind_rows(risk_parts) else tibble(
  marker = character(), cancer_type = character(), cohort = character(), ps_model = character(),
  group = character(), time = double(), n_at_risk = integer(), weighted_at_risk = double()
)
manifest <- if (length(manifest_parts)) bind_rows(manifest_parts) else tibble(
  marker = character(), cancer_type = character(), cohort = character(), ps_model = character(),
  sig_in = character(), HR_markerxICI = double(), p_markerxICI = double(),
  extreme_hr_flag = logical(), status = character(), reason = character(), file = character()
)

readr::write_csv(curves_compiled, file.path(KM_OUT_DIR, "KM_compiled_curves.csv"))
readr::write_csv(at_risk_compiled, file.path(KM_OUT_DIR, "KM_compiled_at_risk.csv"))
readr::write_csv(manifest, file.path(KM_OUT_DIR, "KM_export_manifest.csv"))

if (nrow(manifest)) print(manifest %>% count(cohort, ps_model, status))
cat("Output directory:", KM_OUT_DIR, "\n")
```

## Interpretation

The interaction HR is a ratio of marker hazard ratios under ICI versus non-ICI;
it is not a direct contrast between any two displayed curves. The log-rank values
shown in each panel compare marker-positive with marker-negative patients only
within the ICI arm and are descriptive, especially for the weighted estimate.
Check the at-risk counts and effective sample sizes before interpreting tail
separation or an extreme interaction HR.
